Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-06-07 paper/graph.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@623 3fc257bb-8550-0410-81b3-a5c103da83f7
Lots of small edits + some new XMT results.
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@628 3fc257bb-8550-0410-81b3-a5c103da83f7

$M_{\rm remain} = M$ \\
while $M_{\rm remain} > 0$: \\
\> Map: \> Generate $M_{\rm remain}/P$ random edges on each processor \\
           \> \> output Key = $(i,j)$, Value = NULL \\
\> Collate \\
\> Reduce: \> Remove duplicate edges \\
           \> \> input Key = $(i,j)$, MultiValue = one or more NULLs \\
           \> \> output Key = $(i,j)$, Value = NULL \\
\> $M_{\rm remain} = M - N_{kv}$

@@

 \caption{MapReduce algorithm for R-MAT graph generation.}

@@

The {\it map()} will often generate some small number of duplicate edges.
The {\it collate()} and {\it reduce()} operations remove the duplicates.  The
entire map-collate-reduce sequence is repeated until the number of
resulting key/value pairs $N_{kv} = M$.  For reasonably sparse graphs
this typically takes only a few iterations.

@@

By using more than one MapReduce object, we can improve the performance
of the R-MAT generation algorithm, particularly when out-of-core operations
are needed.  In the enhanced algorithm
(Figure~\ref{fig:rmat2}), we emit newly created edges into MapReduce object
$E_{new}$ and {\it aggregate()} the edges to processors.  
We then add those edges
to the MapReduce object $E_{old}$, containing all previously generated
unique edges; the edges of $E_{old}$ are already aggregated to processors
using the same mapping as $E_{new}$.  We can then use the {\it compress()}
function to reduce duplicates locally on each processor.
Time reductions
in this algorithm result from the reduced number of KV pairs communicated in
{\it aggregate()},
compared to the collate operation in the original algorithm, 
particularly in later iterations of the generation where few
edges are created.

@@

$M_{\rm remain} = M$ \\
while $M_{\rm remain} > 0$: \\
\> Map $E_{new}$: \> Generate $M_{\rm remain}/P$ random edges on each processor \\
           \> \> output Key = $(i,j)$, Value = NULL \\
\> Aggregate $E_{new}$. \\
\> Add $E_{new}$ to $E_{old}$. \\
\> Compress $E_{old}$: Remove duplicate edges \\
            \> \> input Key = $(i,j)$, MultiValue = one or more NULLs \\
            \> \> output Key = $(i,j)$, Value = NULL \\
\> $M_{\rm remain} = M - N_{kv}$

@@

 \caption{Enhanced MapReduce algorithm for R-MAT graph generation.}

@@

\subsection{PageRank}

The PageRank algorithm assings a relative numeric rank to each
vertex in a graph.
It models the web as a directed graph $G(V,E)$, with each vertex $V_i
\in V$ representing a web page and each edge $(V_i, V_j) \in E$
representing a hyperlink from $V_i$ to $V_j$.  The probability of
moving from $V_i$ to another vertex $V_j$ is $\alpha/d_{out}(V_i) +
(1-\alpha)/|V|$, where $\alpha$ is a user-defined parameter (usually
0.8-0.9), $d_{out}(V_i)$ is the outdegree of vertex $V_i$, and $|V|$ is
the cardinality of $V$.  The first term represents the probability of
following a given link on page $V_i$; the second represents the
probability of moving to a random page.  For pages with no outlinks,
the first term is $\alpha/|V|$, indicating equal likelihood to move to
any other page.  Equivalently, the graph can be represented by a
matrix $A$~\cite{LangvilleMeyer05a}, with matrix entries $A_{ij} =
\alpha/d_{out}(V_i)$ if vertex $V_i$ links to $V_j$.  To maintain the 
sparsity of $A$, terms for random jumps and zero out-degree vertices are
not explicitly stored in $A$.  Rather, they are computed separately
as adjustments to the PageRank vector.  The kernel of the PageRank
algorithm, then, is a power-method iteration consisting of 
matrix-vector multiplication $A^T x=y$,
where $x$ is the PageRank vector from the previous iteration, and
a few dot product and norm computations.

The algorithm in Figure~\ref{fig:pr2} performs these
iterations.  The nonzeros of matrix $A$ are initialized as described above;
the PageRank vector $x$ is initialized uniformly.  A MapReduce object $MT$
contains the indices of all-zero rows of $A$, corresponding to vertices
with no outlinks.  
The PageRank algorithm is made more efficient by calling
{\it aggregate()} on the MapReduce objects representing the matrix $A$, 
the all-zero matrix rows $MT$, and the vector $x$ once before the PageRank 
iterations begin.  Since these three MapReduce
objects have the same key types, pre-aggregating the objects moves all
Key-Value pairs with a given key to a single processor.  Thus, many of
the PageRank operations become local operations, replacing 
{\it collate()} operations (which is equivalent to {\it aggregate()} 
followed by {\it convert()}) with only local {\it convert()} calls. 
In Figure~\ref{fig:pr2}, steps where {\it convert()} operations can be
substituted for {\it convert()} operations are marked with an asterisk.
This strategy of pre-aggregating to improve data locality is useful in
many MapReduce algorithms, but assumes that keys are always mapped to
the same processor.

In step 1 of a PageRank interation, $MT$ is used 
in dot products to compute adjustments to the
PageRank vector to represent uniform probability of going from a leaf page
to any other page.  Adjustments for random jumps from any page are also
computed in this step.  Step 2, the matrix-vector product $A^T x$, is the most
expensive part of the computation, requiring two passes over the nonzero
structure of $A$.  In the first pass, {\it convert()} 
gathers all local row entries $a_{ij}$ with their
associated $x_i$ entry, and {\it reduce()} computes $a_{ij} x_i$.  
In the second pass, 
{\it collate()} gathers, for each $j$, all contributions to the
column sum $\sum_i a_{ij} x_i$, which is computed by a second {\it reduce()}. 
Steps 3 and 4 adjust and scale the 
product vector; the work is proportional to $|V|$.  Step 6 computes the
residual to determine whether the PageRank vector has converged; work is
again proportional to $|V|$.  Throughout the algorithm, we exploit the
ability to call MPI\_Allreduce to compute global norms and residuals.

%\begin{figure}[htb]
% \begin{center}\fbox{\begin{minipage}{\textwidth} \begin{tabbing}
%  xxxx\=xxxxx\=xxxxx\=xxxxx\=xxxxxxxxxxxxxxxxxxxxxxxxxxx\= \kill
%Input: \\
%\> MapReduce object $A$ containing $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, a_{ij}]$ \\
%\> MapReduce object $MT$ containing indices of all-zero rows of $A$:  Key = $i$, Value = 0. \\
%\> MapReduce object $x$ containing initial PageRank vector:  Key = $i$, Value = $1/|V|$. \\
%While (${residual} > {tolerance}$) \\
%\> 1. Compute contribution for random jumps and zero-outdegree vertices. \\
%\> \> Add: $x$ to $MT$. \\
%\> \> Collate $MT$:  Vertex is Key. \\
%\> \> Reduce $MT$:   \\
%\> \> \> Input:  Key = $i$; MultiValue = $x_i, 0$ if $i \in {MT}$; MultiValue = $x_i$ if $i \notin {MT}$. \\
%\> \> \> Compute:  Local contributions $c_{local}$. \\
%\> \> \> Output:  If ${nvalues} = 2$, Key = $i$; Value = 0. \\
%\> \> MPI\_Allreduce $c_{local}$ with MPI\_SUM to compute global contribution $c_{global}$. \\
%\> 2. Compute $y = A^T x$.  \\
%\> \> Copy $x$ to $y$. \\
%\> \> Add $A$ to $y$. \\
%\> \> Collate $y$:  Key is row index of $A$. \\
%\> \> Reduce $y$: \\
%\> \> \>  Input: Key = $i$; Multivalue = $i, [j, a_{ij}]$ for nonzeros $a_{ij}$ in $A$ \\
%\> \> \>  Output:  Key = $j$; Value = $a_{ij} y_i$. \\
%\> \> Collate $y$:  Key is column index of $A$. \\
%\> \> Reduce $y$:  \\
%\> \> \> Input:  Key = $j$; Multivalue = $a_{ij} y_i$ for all non-zero entries $a_{ij}$ of column $j$ \\
%\> \> \> Output:  Key = $j$; Value = $\sum_i a_{ij} y_i$. \\
%\> 3. Add contribution $c_{global}$ to $y$; compute local max norm $n_{local}$ of $y$. \\
%\> \> Map $y$:  \\
%\> \> \> Input:  Key = $i$, Value = $y_i$. \\
%\> \> \> Compute: $y_i = y_i + c_{global}$; if $y_i > n_{local}$, $n_{local} = y_i$. \\
%\> \> \> Output:  Key = $i$, Value = $y_i$. \\
%\> \> MPI\_Allreduce $n_{local}$ with MPI\_MAX to compute global max norm $n_{global}$. \\
%\> 4. Scale $y$ by $n_{global}$. \\
%\> \> Map $y$: \\
%\> \> \> Input:  Key = $i$, Value = $y_i$. \\
%\> \> \> Output:  Key = $i$, Value = $y_i / {n_{global}}$. \\
%\> 5. Compute residual. \\
%\> \> Add $y$ to $x$. \\
%\> \> Collate $x$:  Key is index $i$. \\
%\> \> Reduce $x$:   \\
%\> \> \> Input:  Key = $i$; Multivalue = $x_i, y_i$. \\
%\> \> \> Compute:  if $|x_i - y_i| > {resid}$, ${resid} = |x_i - y_i|$ \\
%\> \> MPI\_Allreduce with MPI\_MAX to compute $residual$. \\
%\> 6. $x = y$; \\
%Output:  PageRank vector $x$.
  %\end{tabbing}
 %\end{minipage}}\end{center}
%
 %\caption{MapReduce algorithm for PageRank vertex ranking.}
%
 %\label{fig:pr}
%\end{figure}

@@

Input: \\
$*$\> Pre-aggregated MapReduce object $A$ containing $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, a_{ij}]$ \\
$*$\> Pre-aggregated MapReduce object $MT$ containing indices of all-zero rows of $A$:  Key = $i$, Value = 0. \\
$*$\> Pre-aggregated MapReduce object $x$ containing initial PageRank vector:  Key = $i$, Value = $1/|V|$. \\
While (${residual} > {tolerance}$) \\
\> Compute contribution for random jumps and zero-outdegree vertices. \\
\> \> Add: $x$ to $MT$. \\
$*$\> \> Convert $MT$:  Vertex is Key. \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$; MultiValue = $x_i, 0$ if $i \in {MT}$; MultiValue = $x_i$ if $i \notin {MT}$. \\
\> \> \> Compute:  Local contributions $c_{local}$. \\
\> \> \> Output:  If ${nvalues} = 2$, Key = $i$; Value = 0. \\
\> \> MPI\_Allreduce $c_{local}$ with MPI\_SUM to compute global contribution $c_{global}$. \\
\> Compute $y = A^T x$.  \\
\> \> Copy $x$ to $y$. \\
\> \> Add $A$ to $y$. \\
$*$\> \> Convert $y$:  Key is row index of $A$. \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$; Multivalue = $i, [j, a_{ij}]$ for nonzeros $a_{ij}$ in $A$ \\
\> \> \>  Output:  Key = $j$; Value = $a_{ij} y_i$. \\
\> \> Collate $y$:  Key is column index of $A$. \\
\> \> Reduce $y$:  \\
\> \> \> Input:  Key = $j$; Multivalue = $a_{ij} y_i$ for all non-zero entries $a_{ij}$ of column $j$ \\
\> \> \> Output:  Key = $j$; Value = $\sum_i a_{ij} y_i$. \\
\> Add contribution $c_{global}$ to $y$; compute local max norm $n_{local}$ of $y$. \\
\> \> Map $y$:  \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Compute: $y_i = y_i + c_{global}$; if $y_i > {n_{local}}$, $n_{local} = y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i$. \\
\> \> MPI\_Allreduce $n_{local}$ with MPI\_MAX to compute global max norm $n_{global}$. \\
\> Scale $y$ by $n_{global}$. \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i / {n_{global}}$. \\
\> Compute residual. \\
\> \> Add $y$ to $x$. \\
$*$\> \> Convert $x$:  Key is index $i$. \\
\> \> Reduce $x$:   \\
\> \> \> Input:  Key = $i$; Multivalue = $x_i, y_i$. \\
\> \> \> Compute:  if $|x_i - y_i| > {resid}$, ${resid} = |x_i - y_i|$ \\
\> \> MPI\_Allreduce with MPI\_MAX to compute $residual$. \\
\> $x = y$; \\
Output:  PageRank vector $x$.
  \end{tabbing}
 \end{minipage}}\end{center}

 \caption{MapReduce algorithm for PageRank vertex ranking; pre-aggregating the MapReduce objects allows many communication intensive {\it collate()} 
operations to be replaced by with strictly local {\it convert()} operations (as marked by the asterisks).}

@@

In stage 4, the KV pairs in the original graph $G_0$ are added to the
current working set of KV pairs.  The KV pairs in $G_0$ are edges that
complete triangles for the angle KV pairs just generated.  After the
fourth {\it collate()}, a pair of vertices $(V_i,V_j)$ is the key, and
the multivalue is the list of all root vertices in angles that contain
$V_i$ and $V_j$.  If the multivalue also contains a NULL, contributed
by $G_0$, then there is a $(V_i,V_j)$ edge in the graph.  Thus all
vertices in the multivalue are roots of angles which are complete
triangles and can be emitted as a triplet key.

@@

Iterate: \\
\> 1 Map: \> Convert edges to vertices \\
    \> \> \> input Key = $E_{ij}$, Value = NULL \\
    \> \> \> output Key = $V_i$, Value = $E_{ij}$ \\
    \> \> \> output Key = $V_j$, Value = $E_{ij}$ \\
\> 1 Add: \> Zone assignment of each vertex \\
    \> \> \> output Key = $V_i$, Value = $Z_i$ \\
\> 1 Collate: \> Vertex as key \\
\> 1 Reduce: \> Emit edges of each vertex with zone of vertex \\
       \> \> \> input Key = $V_i$, MultiValue = $E E E E ... Z$ \\
       \> \> \> for each $E$ in MultiValue: \\
      \> \> \> \> output Key = $E_{ij}$, Value = $Z_i$ \\

@@

The algorithm begins (before the iteration loop) by assigning each
vertex to its own component or ``zone,'' so that $Z_i$ = $V_i$.  Each
iteration will grow the zones, one layer of neighbors at a time.  As
zones collide due to shared edges, a winner is chosen (the smaller
zone ID), and vertices in the losing zone are reassigned to the
winning zone.  When the iterations complete, each zone will have
become a fully connected component.  The algorithm thus finds all
connected components in the graph simultaneously.  The number of
iterations required depends on the largest diameter of any component
in the graph.

The first {\it map()} operation emits the vertices in each edge as
keys, with the edge as a value.  The current zone assignment of each
vertex is added to the set of key/value pairs.  The first {\it
collate()} operation collects all the edges of a vertex and its zone
assignment together in one multi-value.  The first {\it reduce()}
operation then re-emits each edge, tagged by the zone assignment of
one of its vertices.

@@

The third {\it map()} operation inverts the vertex/zone key/value
pairs to become zone/vertex pairs.  The third {\it add()} operation
adds the changing zone assignments to the set of key/value pairs.  The
third collate() can then collect all the vertices of a zone and zero
or more reassignments for the zone ID.  Since a zone could collide
with multiple other zones on the same iteration due to shared edges,
the new zone ID becomes the minimum ID of any of the neighboring
zones.  If no zone reassignment value appears in the multi-value, the
zone ID is unchanged.  The final reduce() a key/value pair for each
vertex in the zone, with the vertex as a key and the new zone ID as a
value.

Note that if a graph has only a few components, the third {\it
collate()} operation, which collates by the zone ID, may generate a few
very large key/multi-value (KMV) pairs.  For example, if the graph is
fully connected, then on the last iteration, a single KMV pair will
contain all vertices in the graph and be assigned to one processor.
This imbalance in memory and computational work can lead to poor
parallel performance of the overall algorithm.  To counter this
effect, the various operations of stage 3 include extra logic.  The
idea is to partition zones whose vertex count exceeds a user-defined
threshhold into $P$ sub-zones, where $P$ is the number of processors.
The 64-bit integer that stores the zone ID also stores a bit flag
indicating the zone has been partitioned and a set of bits that encode
the processor ID.

During the third {\it map()} operation, if the zone has been
partitioned, then the vertex is assigned to a random processor and the
processor ID bits are added to the zone ID, as indicated by the
$Z_i^+$ notation in Figure~\ref{fig:cc}.  Likewise, if $Z_i$ has been
partitioned, the third {\it add()} operation emits the zone ID change
key/value pair ($Z_i$,$Z_{winner}$) as ($Z_i^+$,$Z_{winner}$).  In
this case ($Z_i$,$Z_{winner}$), is emitted not once, but $P+1$ times,
once for each processor, and once as if $Z_i$ had not been partitioned
(for a reason discussed below).

This additional partitioning logic means that the third {\it
collate()} operation, which keys on zone IDs, some of which now
include processor bits, will collect only a $1/P$ subset of the
vertices in large zones onto each processor.  But the multivalue on
each processor contains all the zone-reassignments relevant to the
unpartitioned zone.  This allows the third {\it reduce()} operation to
change the zone ID (if necessary) in a consistent manner across all
$P$ multi-values that contain the zone's vertices.  When the {\it
reduce()} operation emits new zone assignments for each vertex, the
zone retains its partitioned status, and the partition bit is also
explicitly set if the vertex count exceeds the threshhold for the
first time.

>>>>

$M_{\rm remain} = M$ \\
while $M_{\rm remain} > 0$: \\
\> Map: \> Generate $M_{\rm remain}/P$ random edges $(i,j)$ on each processor \\
           \> \> output Key = $(i,j)$, Value = NULL \\
\> Collate \\
\> Reduce: \> Remove duplicate edges \\
           \> \> input Key = $(i,j)$, MultiValue = one or more NULLs \\
           \> \> output Key = $(i,j)$, Value = NULL \\
\> $M_{\rm remain} = M - N_{kv}$

@@

 \caption{MapReduce algorithm for R-MAT graph generation on $P$ processors.}

@@

The {\it map()} often generates some small number of duplicate edges.
The {\it collate()} and {\it reduce()} operations remove the duplicates.  The
entire map-collate-reduce sequence is repeated until the number of
resulting key/value pairs $N_{kv} = M$.  For reasonably sparse graphs
this typically takes only a few iterations.

@@

By using more than one MapReduce object, we can improve the performance
of the R-MAT generation algorithm, particularly when out-of-core operations
are needed.  In the enhanced algorithm
(Figure~\ref{fig:rmat2}), we emit newly created edges into MapReduce object
$E_{new}$ and {\it aggregate()} the edges to processors.  
We then add those edges
to the MapReduce object $E_{old}$, containing all previously generated
unique edges; the edges of $E_{old}$ are already aggregated to processors
using the same mapping as $E_{new}$.  We can then use the {\it compress()}
function to reduce duplicates locally on each processor.
Time reductions
in this algorithm result from the smaller number of KV pairs communicated in
{\it aggregate()},
compared to the collate operation in the original algorithm, 
particularly in later iterations of the generation where few
edges are created.

@@

$M_{\rm remain} = M$ \\
while $M_{\rm remain} > 0$: \\
\> Map $E_{new}$: \> Generate $M_{\rm remain}/P$ random edges $(i,j)$ on each processor \\
           \> \> output Key = $(i,j)$, Value = NULL \\
\> Aggregate $E_{new}$. \\
\> Add $E_{new}$ to $E_{old}$. \\
\> Compress $E_{old}$: Remove duplicate edges \\
            \> \> input Key = $(i,j)$, MultiValue = one or more NULLs \\
            \> \> output Key = $(i,j)$, Value = NULL \\
\> $M_{\rm remain} = M - N_{kv}$

@@

 \caption{Enhanced MapReduce algorithm for R-MAT graph generation on $P$ processors.}

@@

\subsection{PageRank}
\label{subsec:graph_pagerank}

The PageRank algorithm assigns a relative numeric rank to each
vertex in a graph.
It models the web as a directed graph $G(V,E)$, with each vertex $V_i
\in V$ representing a web page and each edge $(V_i, V_j) \in E$
representing a hyperlink from $V_i$ to $V_j$.  The probability of
moving from $V_i$ to another vertex $V_j$ is $\alpha/d_{out}(V_i) +
(1-\alpha)/|V|$, where $\alpha$ is a user-defined parameter (usually
0.8-0.9), $d_{out}(V_i)$ is the outdegree of vertex $V_i$, and $|V|$ is
the cardinality of $V$.  The first term represents the probability of
following a given link on page $V_i$; the second represents the
probability of moving to a random page.  For pages with no outlinks,
the first term is $\alpha/|V|$, indicating equal likelihood to move to
any other page.  Equivalently, the graph can be represented by a
matrix $A$~\cite{LangvilleMeyer05a}, with matrix entries $A_{ij} =
\alpha/d_{out}(V_i)$ if vertex $V_i$ links to $V_j$.  To maintain the 
sparsity of $A$, terms for random jumps and zero out-degree vertices are
not explicitly stored in $A$.  Rather, they are computed separately
as adjustments to the PageRank vector.  The kernel of the PageRank
algorithm, then, is a power-method iteration consisting of 
matrix-vector multiplication $A^T x=y$,
where $x$ is the PageRank vector from the previous iteration, and
a few dot product and norm computations.

The algorithm in Figure~\ref{fig:pr2} performs these
iterations.  The nonzeros of matrix $A$ are initialized as described above;
the PageRank vector $x$ is initialized uniformly.  A MapReduce object $MT$
contains the indices of all-zero rows of $A$, corresponding to vertices
with no outlinks.  
The PageRank algorithm is made more efficient by calling
{\it aggregate()} once on the MapReduce objects representing the matrix $A$, 
the all-zero matrix rows $MT$, and the vector $x$ before the PageRank 
iterations begin.  Since these three MapReduce
objects have the same key types, pre-aggregating the objects moves all
Key-Value pairs with a given key to a single processor.  Thus, many of
the PageRank operations become local operations, replacing 
{\it collate()} operations (which are equivalent to {\it aggregate()} 
followed by {\it convert()}) with only local {\it convert()} calls. 
In Figure~\ref{fig:pr2}, steps where {\it convert()} operations can be
substituted for {\it collate()} operations are marked with an asterisk.
This strategy of pre-aggregating to improve data locality is useful in
many MapReduce algorithms, but assumes that keys are always mapped to
the same processor.

In step 1 of a PageRank interation, $MT$ is used 
in dot products to compute adjustments to the
PageRank vector to represent uniform probability of going from a leaf page
to any other page.  Adjustments for random jumps from any page are also
computed in this step.  Step 2, the matrix-vector product $A^T x$, is the most
expensive part of the computation, requiring two passes over the nonzero
structure of $A$.  In the first pass, {\it convert()} 
gathers all local row entries $A_{ij}$ with their
associated $x_i$ entry, and {\it reduce()} computes $A_{ij} x_i$.  
In the second pass, 
{\it collate()} gathers, for each $j$, all contributions to the
column sum $\sum_i A_{ij} x_i$, which is computed by a second {\it reduce()}. 
Steps 3 and 4 adjust and scale the 
product vector; the work is proportional to $|V|$.  Step 6 computes the
residual to determine whether the PageRank vector has converged; work is
again proportional to $|V|$.  Throughout the algorithm, we exploit the
ability to call MPI\_Allreduce to compute global norms and residuals.

%\begin{figure}[htb]
% \begin{center}\fbox{\begin{minipage}{\textwidth} \begin{tabbing}
%  xxxx\=xxxxx\=xxxxx\=xxxxx\=xxxxxxxxxxxxxxxxxxxxxxxxxxx\= \kill
%Input: \\
%\> MapReduce object $A$ containing $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, A_{ij}]$ \\
%\> MapReduce object $MT$ containing indices of all-zero rows of $A$:  Key = $i$, Value = 0. \\
%\> MapReduce object $x$ containing initial PageRank vector:  Key = $i$, Value = $1/|V|$. \\
%While (${residual} > {tolerance}$) \\
%\> 1. Compute contribution for random jumps and zero-outdegree vertices. \\
%\> \> Add: $x$ to $MT$. \\
%\> \> Collate $MT$:  Vertex is Key. \\
%\> \> Reduce $MT$:   \\
%\> \> \> Input:  Key = $i$; MultiValue = $x_i, 0$ if $i \in {MT}$; MultiValue = $x_i$ if $i \notin {MT}$. \\
%\> \> \> Compute:  Local contributions $c_{local}$. \\
%\> \> \> Output:  If ${nvalues} = 2$, Key = $i$; Value = 0. \\
%\> \> MPI\_Allreduce $c_{local}$ with MPI\_SUM to compute global contribution $c_{global}$. \\
%\> 2. Compute $y = A^T x$.  \\
%\> \> Copy $x$ to $y$. \\
%\> \> Add $A$ to $y$. \\
%\> \> Collate $y$:  Key is row index of $A$. \\
%\> \> Reduce $y$: \\
%\> \> \>  Input: Key = $i$; Multivalue = $i, [j, A_{ij}]$ for nonzeros $A_{ij}$ in $A$ \\
%\> \> \>  Output:  Key = $j$; Value = $A_{ij} y_i$. \\
%\> \> Collate $y$:  Key is column index of $A$. \\
%\> \> Reduce $y$:  \\
%\> \> \> Input:  Key = $j$; Multivalue = $A_{ij} y_i$ for all non-zero entries $A_{ij}$ of column $j$ \\
%\> \> \> Output:  Key = $j$; Value = $\sum_i A_{ij} y_i$. \\
%\> 3. Add contribution $c_{global}$ to $y$; compute local max norm $n_{local}$ of $y$. \\
%\> \> Map $y$:  \\
%\> \> \> Input:  Key = $i$, Value = $y_i$. \\
%\> \> \> Compute: $y_i = y_i + c_{global}$; if $y_i > n_{local}$, $n_{local} = y_i$. \\
%\> \> \> Output:  Key = $i$, Value = $y_i$. \\
%\> \> MPI\_Allreduce $n_{local}$ with MPI\_MAX to compute global max norm $n_{global}$. \\
%\> 4. Scale $y$ by $n_{global}$. \\
%\> \> Map $y$: \\
%\> \> \> Input:  Key = $i$, Value = $y_i$. \\
%\> \> \> Output:  Key = $i$, Value = $y_i / {n_{global}}$. \\
%\> 5. Compute residual. \\
%\> \> Add $y$ to $x$. \\
%\> \> Collate $x$:  Key is index $i$. \\
%\> \> Reduce $x$:   \\
%\> \> \> Input:  Key = $i$; Multivalue = $x_i, y_i$. \\
%\> \> \> Compute:  if $|x_i - y_i| > {resid}$, ${resid} = |x_i - y_i|$ \\
%\> \> MPI\_Allreduce with MPI\_MAX to compute $residual$. \\
%\> 6. $x = y$; \\
%Output:  PageRank vector $x$.
  %\end{tabbing}
 %\end{minipage}}\end{center}
%
 %\caption{MapReduce algorithm for PageRank vertex ranking.}
%
 %\label{fig:pr}
%\end{figure}

@@

Input: \\
$*$\> Pre-aggregated MapReduce object $A$ containing $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, A_{ij}]$ \\
$*$\> Pre-aggregated MapReduce object $MT$ containing indices of all-zero rows of $A$:  Key = $i$, Value = 0. \\
$*$\> Pre-aggregated MapReduce object $x$ containing initial PageRank vector:  Key = $i$, Value = $1/|V|$. \\
While (${residual} > {tolerance}$) \\
\> Compute contribution for random jumps and zero-outdegree vertices. \\
\> \> Add: $x$ to $MT$. \\
$*$\> \> Convert $MT$:  Vertex is Key. \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$; MultiValue = $x_i, 0$ if $i \in {MT}$; MultiValue = $x_i$ if $i \notin {MT}$. \\
\> \> \> Compute:  Local contributions $c_{local}$. \\
\> \> \> Output:  If ${nvalues} = 2$, Key = $i$; Value = 0. \\
\> \> MPI\_Allreduce $c_{local}$ with MPI\_SUM to compute global contribution $c_{global}$. \\
\> Compute $y = A^T x$.  \\
\> \> Copy $x$ to $y$. \\
\> \> Add $A$ to $y$. \\
$*$\> \> Convert $y$:  Key is row index of $A$. \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$; Multivalue = $i, [j, A_{ij}]$ for nonzeros $A_{ij}$ in $A$ \\
\> \> \>  Output:  Key = $j$; Value = $A_{ij} y_i$. \\
\> \> Collate $y$:  Key is column index of $A$. \\
\> \> Reduce $y$:  \\
\> \> \> Input:  Key = $j$; Multivalue = $A_{ij} y_i$ for all non-zero entries $A_{ij}$ of column $j$ \\
\> \> \> Output:  Key = $j$; Value = $\sum_i A_{ij} y_i$. \\
\> Add contribution $c_{global}$ to $y$; compute local max norm $n_{local}$ of $y$. \\
\> \> Map $y$:  \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Compute: $y_i = y_i + c_{global}$; if $y_i > {n_{local}}$, $n_{local} = y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i$. \\
\> \> MPI\_Allreduce $n_{local}$ with MPI\_MAX to compute global max norm $n_{global}$. \\
\> Scale $y$ by $n_{global}$. \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$. \\
\> \> \> Output:  Key = $i$, Value = $y_i / {n_{global}}$. \\
\> Compute residual. \\
\> \> Add $y$ to $x$. \\
$*$\> \> Convert $x$:  Key is index $i$. \\
\> \> Reduce $x$:   \\
\> \> \> Input:  Key = $i$; Multivalue = $x_i, y_i$. \\
\> \> \> Compute:  if $|x_i - y_i| > {resid}$, ${resid} = |x_i - y_i|$ \\
\> \> MPI\_Allreduce with MPI\_MAX to compute $residual$. \\
\> $x = y$; \\
Output:  PageRank vector $x$.
  \end{tabbing}
 \end{minipage}}\end{center}

 \caption{MapReduce algorithm for PageRank vertex ranking; pre-aggregating the MapReduce objects allows many communication intensive {\it collate()} 
operations to be replaced by strictly local {\it convert()} operations (as marked by the asterisks).}

@@

In stage 4, the KV pairs in the original graph $G_0$ are added to the
current working set of KV pairs.  The KV pairs in $G_0$ are edges that
complete triangles for the angle KV pairs just generated.  After the
fourth {\it collate()}, a pair of vertices $(V_i,V_j)$ is the key, and
the multivalue is the list of all root vertices in angles that contain
$V_i$ and $V_j$.  If the multivalue also contains a NULL, contributed
by $G_0$, then there is a $(V_i,V_j)$ edge in the graph.  Thus all
vertices in the multivalue are roots of angles that are complete
triangles and can be emitted as a triplet key.

@@

Iterate: \\
\> 1 Map: \> Convert edges to vertices \\
    \> \> \> input Key = $E_{ij} = (V_i, V_j)$, Value = NULL \\
    \> \> \> output Key = $V_i$, Value = $E_{ij}$ \\
    \> \> \> output Key = $V_j$, Value = $E_{ij}$ \\
\> 1 Add: \> Zone assignment of each vertex \\
    \> \> \> output Key = $V_i$, Value = $Z_i$ \\
\> 1 Collate: \> Vertex as key \\
\> 1 Reduce: \> Emit edges of each vertex with zone of vertex \\
       \> \> \> input Key = $V_i$, MultiValue = $E E E E ... Z$ \\
       \> \> \> for each $E$ in MultiValue: \\
      \> \> \> \> output Key = $E_{ij}$, Value = $Z_i$ \\

@@

The algorithm begins (before the iteration loop) by assigning each
vertex to its own component or ``zone,'' so that $Z_i$ = $V_i$.  Each
iteration grows the zones, one layer of neighbors at a time.  As
zones collide due to shared edges, a winner is chosen (the smaller
zone ID), and vertices in the losing zone are reassigned to the
winning zone.  When the iterations complete, each zone has
become a fully connected component.  The algorithm thus finds all
connected components in the graph simultaneously.  The number of
iterations required depends on the largest diameter of any component
in the graph.

The first {\it map()} operation emits the vertices in each edge as
keys, with the edge as a value.  The current zone assignment of each
vertex is added to the set of key/value pairs.  The first {\it
collate()} operation collects all the edges of a vertex and its zone
assignment together in one multi-value.  The first {\it reduce()}
operation re-emits each edge, tagged by the zone assignment of
one of its vertices.

@@

The third {\it map()} operation inverts the vertex/zone key/value
pairs to become zone/vertex pairs.  The third {\it add()} operation
adds the changing zone assignments to the set of key/value pairs.  The
third {\it collate()} can then collect all the vertices of a zone and zero
or more reassignments for the zone ID.  Since a zone could collide
with multiple other zones on the same iteration due to shared edges,
the new zone ID becomes the minimum ID of any of the neighboring
zones.  If no zone reassignment value appears in the multi-value, the
zone ID is unchanged.  The final {\it reduce()} emits a key/value pair for each
vertex in the zone, with the vertex as a key and the new zone ID as a
value.

Note that if a graph has only a few components, the third {\it
collate()} operation, which collates by the zone ID, may generate a few
very large key/multi-value (KMV) pairs.  For example, if the graph is
fully connected, in the last iteration, a single KMV pair 
contains all vertices in the graph and is assigned to one processor.
This imbalance in memory and computational work can lead to poor
parallel performance of the overall algorithm.  To counter this
effect, the various operations of stage 3 include extra logic.  The
idea is to partition zones whose vertex count exceeds a user-defined
threshhold into $P$ sub-zones, where $P$ is the number of processors.
The 64-bit integer that stores the zone ID also stores a bit flag
indicating the zone has been partitioned and a set of bits that encode
the processor ID.

During the third {\it map()} operation, if the zone has been
partitioned, the vertex is assigned to a random processor and the
processor ID bits are added to the zone ID, as indicated by the
$Z_i^+$ notation in Figure~\ref{fig:cc}.  Likewise, if $Z_i$ has been
partitioned, the third {\it add()} operation emits the zone ID change
key/value pair ($Z_i$,$Z_{winner}$) as ($Z_i^+$,$Z_{winner}$).  In
this case ($Z_i$,$Z_{winner}$), is emitted not once, but $P+1$ times,
once for each processor, and once as if $Z_i$ had not been partitioned
(for a reason discussed below).

With this additional partitioning logic, the third {\it
collate()} operation (which keys on zone IDs, some of which now
include processor bits) collects only $1/P$ of the
vertices in large zones onto each processor.  But the multivalue on
each processor contains all the zone-reassignments relevant to the
unpartitioned zone.  Thus, the third {\it reduce()} operation can
change the zone ID (if necessary) in a consistent manner across all
$P$ multi-values that contain the zone's vertices.  When the {\it
reduce()} operation emits new zone assignments for each vertex, the
zone retains its partitioned status; the partition bit is also
explicitly set if the vertex count exceeds the threshhold for the
first time.
